Antisense transcripts can quickly evolve to encode proteins

### 2023-03-10 Manuscripts/M2/antisense_genes.tex
new file:   Manuscripts/Figures/M2_main/pORF_antisense_dmel.svg 	modified:   Manuscripts/Figures/M2_main/pORF_antisense_scer.svg 	modified:   Manuscripts/M2/antisense_genes.tex 	modified:   Modeling/antisenseGenes.jl

\renewenvironment{abstract}{%
  \normalsize
  \textbf{\large\abstractname\\[1em]} % with a normal space
}

@@

\onehalfspacing
\def\figdir{../Figures/M1_main}

@@

\subsection*{Antisense ORFs are more likely exist in frame 1}

@@

@@

GC content may not accurately describe the nucleotide composition of a locus. Therefore, we also calculated the probability of antisense ORFs using actual codon and dicodon frequencies in yeast ORFs. Likewise, we calculated the probability of intergenic ORFs using the frequencies of DNA trimers in yeast intergenic genome. With this analysis, we found that antisense ORFs longer than 54 and 75 codons in frames 1 and 2, respectively, are more likely to occur than intergenic ORF of the same lengths. For all ORF lengths, antisense ORFs in frame 3 are less likely to occur than intergenic ORFs.

>>>>

@@

\onehalfspacing
\def\figdir{../Figures/M2_main/pdf}

@@

\subsection*{Antisense ORFs are more likely to exist in frame 1}

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{\figdir/pORF_antisense_scer.pdf}
\caption{Antisense ORFs in frame 1 are more likely to exist than intergenic ORFs of identical lengths. Panel \textbf{(A)} shows the probability of antisense ORFs in frame 1 relative to that of intergenic ORFs ($\log_2$ ratio, vertical axis). Line colors indicate the GC-content of the ORFs (yellow = 30\%, blue = 40\%, red = 50\%, purple = 60\%). We do not show antisense ORFs in frames 0 and 2 because their probabilities are identical to that of intergenic ORFs. Panel \textbf{(B)} shows the probability of antisense ORFs relative to that of intergenic ORFs ($\log_2$ ratio, vertical axis), calculated using frequencies of short DNA sequences from yeast genome. Frames 0, 1 and 2 are denoted by dotted, solid and dashed lines, respectively. Horizontal axis in both panels shows the length of the ORFs. For data in both panels, we assume that antisense ORFs overlap completely with the sense ORF.}
\label{pORF}
\end{figure}

@@

GC content may not accurately describe the nucleotide composition of a locus. Therefore, we also calculated the probability of antisense ORFs using actual codon and dicodon frequencies in yeast ORFs. Likewise, we calculated the probability of intergenic ORFs using the frequencies of DNA trimers in yeast intergenic genome. With this analysis, we found that antisense ORFs longer than 54 and 75 codons in frames 1 and 2, respectively, are more likely to exist than intergenic ORF of the same lengths. For all ORF lengths, antisense ORFs in frame 3 are less likely to exist than intergenic ORFs.

Our complementary analyses using \textit{D. melanogaster} as a model for mutation bias, produced similar results. However, when we computed the ORF probabilities using the frequencies of codons, dicodons and intergenic trimers from  \textit{D. melanogaster}, we found that frame 0 was most likely to harbor an antisense ORF.

\subsection*{Antisense ORFs are more likely to emerge in, and less likely to be lost from frame 1}

We next analysed how likely it is for antisense ORFs to emerge, when they are not already present. To this end, we calculated gain probability of antisense ORFs in each of the three frames, and under three different intensities of purifying selection. We also calculated the probability of ORF gain in the intergenic regions. We found that antisense ORFs are less likely to emerge in frames 0 and 2, than ORFs in intergenic regions, for all ORF lengths and GC-content. With increasing intensity of purifying selection, the emergence of these antisense ORFs became increasingly less likely than that of intergenic ORFs.
